State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-04-15 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
subfigs
updated estimation section

The state estimation technique utilises the Unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010}. The URTSS uses the Unscented Kalman Filter (UKF)~\cite{Julier1997} in the forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward path to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. An approximation of the state density is made by propagating the so-called sigma points through the state-space model. The sigma points describe the mean and covariance of the state estimate and are calculated using the unscented transform. The sigma vectors $\mathcal X_i$ are calculated by
\begin{equation}
 \mathcal X_{0}=\bar x
\end{equation}
\begin{equation}
 \mathcal X_{i}=\bar x+(\sqrt{( L + \lambda)\mathbf P_x})_i \quad i=1, \dots,  L
\end{equation}
\begin{equation}
 \mathcal X_{i}=\bar x-(\sqrt{( L + \lambda)\mathbf P_x})_{i- L} \quad i= L+1, \dots, 2 L
\end{equation}
where $L$ is the dimension of the state space, the number of sigma points required is $2L+1$. The scaling parameter $\lambda$ is defined as a scaling parameter
\begin{equation}
\lambda=\alpha^2( L+\kappa)- L 
\end{equation}
where $\alpha$, $\beta$ and $\kappa$ are constants (see ~\cite{Haykin2001} for more details). It should be noted that $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$ and $\mathbf{P}_x$ represents the corresponding covariance matrix from the filtering or smoothing respectively. The sigma vectors are propagated through the nonlinear function The weights of the columns of the sigma point matrix and their corresponding weights used in the UKF are calculated by

\begin{equation}
 \mathbf W_0^{(m)}=\frac{\lambda}{ L+\lambda}
\end{equation}
\begin{equation}
 \mathbf W_0^{(c)}=\frac{\lambda}{ L+\lambda}+(1-\alpha^2+\beta)
\end{equation}
\begin{equation}
 \mathbf W_i^{(m)}=\mathbf W_i^{(c)}=\frac{1}{2( L+\lambda)} \quad i=1, \dots, 2L
\end{equation}
% For explanation of $\lambda, \quad \alpha\quad\text{and}\quad\beta$ see \cite{wan2000unscented,Haykin2001}. $ (\sqrt{(\mathbf L + \lambda)\mathbf P_x})_i$ is the ith column of the matrix square root (e.g., lower-triangular Cholesky factorization). These sigma points propogated through the nonlinear function

 The accuracy of the approximation is linked to the time step used in the temporal discretization of the system, where the state transition appears more linear with a finer resolution.

To explain the operation of the UKF we begin by deriving the traditional Kalman filter with the state-space model
\begin{eqnarray}
	\mathbf{x}_t = q(\mathbf{x}_t)\boldsymbol{\theta} - \xi\mathbf{x}_t + k\left(u_t\left(\mathbf{r}\right)\right) + \mathbf{e}_t \\
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,
\end{eqnarray}
where $\boldsymbol{\theta}$ and $\xi$ are the systems parameters. In forwards iteration since the observation equation is linear in parameters we only need to use Unsented transformation in prediction step. The update step equations remains the same as the standard Kalman filtering formulations. In backward smoothing path, estimation from the forward filtering is used to form the new set of sigma points in a similar way to the filtering step, however the predicted cross-covariance of the states need to be calculated. A summary of the algorithm is given in Algorithm 
\begin{algorithm}
\caption{The Unscented Kalman Filter}\label{CPTEMDAlgorithm}
\begin{algorithmic}[1]
\State Initialise
\begin{equation*}
\hat{\mathbf x}_0, \mathbf P_0
\end{equation*}
\State Forward iteration
For $t \in \left\lbrace 0,\cdots, N\right\rbrace $,
\begin{equation*}
\mathcal X_t^f=[\hat{\mathbf x}_t^f \quad \hat{\mathbf x}_t^f+\gamma\sqrt{\mathbf P_t^f} \quad \hat{\mathbf x}_t^{f}-\gamma\sqrt{\mathbf P_t^f}]
\end{equation*}
Propagate the sigma points through the state equation
\begin{equation}
\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_t^f,u_t)
\end{equation}
\begin{equation*}
 \hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-}
\end{equation*}
\begin{equation*}
 \mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e
\end{equation*}
\begin{equation*}
\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_v)^{-1}
\end{equation*}
\begin{equation*}
\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-})
\end{equation*}
\begin{equation*}
 \mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-}
\end{equation*}
\State Backward iteration
\State initialisation
\begin{equation}
 \mathbf P_T^b= \mathbf P_T^f
\end{equation}
\begin{equation}
\hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T
\end{equation}
For $t \in \left\lbrace T-1, \cdots, 0 \right\rbrace $
\begin{equation*}
\mathcal X_t^b=[\hat{\mathbf x}_t^f \quad \hat{\mathbf x}_t^f+\gamma\sqrt{\mathbf P_t^f} \quad \hat{\mathbf x}_t^f-\gamma\sqrt{\mathbf P_t^f}]
\end{equation*}
\State Calculate a new set of  sigma points, $\mathcal X_{i,t+1}^{b-}$ similar to the forward step and compute
\begin{equation*}
 \hat{\mathbf x}_{t+1}^{b-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{b-}
\end{equation*}
\begin{equation*}
 \mathbf P_{t +1}^{b-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})^\top+\boldsymbol \Sigma_e
\end{equation*}
\begin{equation*}
 \mathbf M_{t +1}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t}^{b-}-\hat{\mathbf x}_{t}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t+1}^{f})^\top
\end{equation*}
\State compute the smoother gain $\mathbf S_t$, the smoothed state $\hat{\mathbf x}_t^b $ and the covariance $\mathbf P_{t}^{b} $
\begin{equation*}
\mathbf S_t=\mathbf M_{t +1}\left[ \mathbf P_{t +1}^{b-}\right] ^{-1}
\end{equation*}
\begin{equation*}
 \hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathbf S_t\left[\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{b-}\right] 
\end{equation*}
\begin{equation*}
 \mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathbf S_t\left[\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{b-} \right]\mathbf S_t^\top 
\end{equation*}
\end{algorithmic}
\end{algorithm}

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation utilises a least squares (LS) method that minimises the sum of the squared errors (of a predicted state update) with each new state estimate (see Appendix~\ref{LeastSquaresAppendix}).

Alternatives to the UKF are the Extended Kalman Filter (EKF)~\cite{Haykin2001} and the Sequential Monte Carlo (SMC) filter~\cite{doucet2001}. The EKF approximates the state transition equation by linearizing about the current state estimates (by calculating the Jacobian) at each time instance. The linearisation maintains the Gaussianality of the model. The unscented Kalman filter approximates the \emph{a posteriori} state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, whilst maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, as the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the
EKF can be problematic. SMC filtering can theoretically provide an exact \emph{a posteriori} state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to computation demands. The method is appropriate for state estimation of stochastic nonlinear dynamical systems.

>>>>

The state estimation technique utilises the Unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010}. The URTSS uses the Unscented Kalman Filter (UKF)~\cite{Julier1997} in the forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward path to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. An approximation of the state density is made by propagating the so-called sigma points through the state-space model. The sigma points describe the mean and covariance of the state estimate and are calculated using the unscented transform. The sigma vectors $\mathcal X_i$ are calculated by
\begin{equation}
 \mathcal X_{0}=\bar x
\end{equation}
\begin{equation}
 \mathcal X_{i}=\bar x+(\sqrt{( L + \lambda)\mathbf P_x})_i \quad i=1, \dots,  L
\end{equation}
\begin{equation}
 \mathcal X_{i}=\bar x-(\sqrt{( L + \lambda)\mathbf P_x})_{i- L} \quad i= L+1, \dots, 2 L
\end{equation}
where $L$ is the dimension of the state space, the number of sigma points required is $2L+1$. The scaling parameter $\lambda$ is defined as a scaling parameter
\begin{equation}
\lambda=\alpha^2( L+\kappa)- L 
\end{equation}
where $\alpha$, $\beta$ and $\kappa$ are constants (see ~\cite{Haykin2001} for more details). It should be noted that $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$ and $\mathbf{P}_x$ represents the corresponding covariance matrix from the filtering or smoothing respectively. The sigma vectors are propagated through the system equations and weighted to form the posterior mean and covariance. The weights are calculated by
\begin{equation}
 \mathbf W_0^{(m)}=\frac{\lambda}{ L+\lambda}
\end{equation}
\begin{equation}
 \mathbf W_0^{(c)}=\frac{\lambda}{ L+\lambda}+(1-\alpha^2+\beta)
\end{equation}
\begin{equation}
 \mathbf W_i^{(m)}=\mathbf W_i^{(c)}=\frac{1}{2( L+\lambda)} \quad i=1, \dots, 2L.
\end{equation}
% For explanation of $\lambda, \quad \alpha\quad\text{and}\quad\beta$ see \cite{wan2000unscented,Haykin2001}. $ (\sqrt{(\mathbf L + \lambda)\mathbf P_x})_i$ is the ith column of the matrix square root (e.g., lower-triangular Cholesky factorization). These sigma points propogated through the nonlinear function

The update step equations are the same as the standard Kalman filtering formulations. The state estimates from the forward filtering are used to form a new set of sigma points for the smoother, as described above. However, the predicted cross-covariance of the states needs to be calculated. A summary of the URTSS procedure is given in algorithm~\ref{UKFAlgorithm}.
\begin{algorithm}
\caption{The Unscented Kalman Filter}\label{UKFAlgorithm}
\begin{algorithmic}[1]
\State Initialise
\begin{equation*}
\hat{\mathbf x}_0, \mathbf P_0
\end{equation*}
\State Forward iteration
For $t \in \left\lbrace 0,\cdots, T\right\rbrace $,
\begin{equation*}
\mathcal X_t^f=[\hat{\mathbf x}_t^f \quad \hat{\mathbf x}_t^f+\gamma\sqrt{\mathbf P_t^f} \quad \hat{\mathbf x}_t^{f}-\gamma\sqrt{\mathbf P_t^f}]
\end{equation*}
Propagate the sigma points through the state equation
\begin{equation}
\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_t^f,u_t)
\end{equation}
\begin{equation*}
 \hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-}
\end{equation*}
\begin{equation*}
 \mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e
\end{equation*}
\begin{equation*}
\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_v)^{-1}
\end{equation*}
\begin{equation*}
\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-})
\end{equation*}
\begin{equation*}
 \mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-}
\end{equation*}
\State Backward iteration
\State initialisation
\begin{equation}
 \mathbf P_T^b= \mathbf P_T^f
\end{equation}
\begin{equation}
\hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T
\end{equation}
For $t \in \left\lbrace T-1, \cdots, 0 \right\rbrace $
\begin{equation*}
\mathcal X_t^b=[\hat{\mathbf x}_t^f \quad \hat{\mathbf x}_t^f+\gamma\sqrt{\mathbf P_t^f} \quad \hat{\mathbf x}_t^f-\gamma\sqrt{\mathbf P_t^f}]
\end{equation*}
\State Calculate a new set of  sigma points, $\mathcal X_{i,t+1}^{b-}$ similar to the forward step and compute
\begin{equation*}
 \hat{\mathbf x}_{t+1}^{b-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{b-}
\end{equation*}
\begin{equation*}
 \mathbf P_{t +1}^{b-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})^\top+\boldsymbol \Sigma_e
\end{equation*}
\begin{equation*}
 \mathbf M_{t +1}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t}^{b-}-\hat{\mathbf x}_{t}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t+1}^{f})^\top
\end{equation*}
\State compute the smoother gain $\mathbf S_t$, the smoothed state $\hat{\mathbf x}_t^b $ and the covariance $\mathbf P_{t}^{b} $
\begin{equation*}
\mathbf S_t=\mathbf M_{t +1}\left[ \mathbf P_{t +1}^{b-}\right] ^{-1}
\end{equation*}
\begin{equation*}
 \hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathbf S_t\left[\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{b-}\right] 
\end{equation*}
\begin{equation*}
 \mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathbf S_t\left[\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{b-} \right]\mathbf S_t^\top 
\end{equation*}
\end{algorithmic}
\end{algorithm}

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation utilises a least squares (LS) method that minimises the sum of the squared errors (of a predicted state update) with each new state estimate (see~\ref{LeastSquaresAppendix}).

% The accuracy of the approximation is linked to the time step used in the temporal discretisation of the system, where the state transition appears more linear with a finer resolution.

Alternatives to the UKF are the Extended Kalman Filter (EKF)~\cite{Haykin2001} and the Sequential Monte Carlo (SMC) filter~\cite{doucet2001}. The EKF approximates the state transition equation by linearizing about the current state estimates (by calculating the Jacobian) at each time instance. The linearisation maintains the Gaussianality of the model. The unscented Kalman filter approximates the \emph{a posteriori} state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, whilst maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, as the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the
EKF can be problematic. SMC filtering can theoretically provide an exact \emph{a posteriori} state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to computation demands. %The method is appropriate for state estimation of stochastic nonlinear dynamical systems.
